\singlespacing \vspace{-1.6cm} \LARGE Trade-offs among plant reproductive traits determine interactions with floral visitors

### 2022-02-01 Manuscript/Lanuza_et_al_2021_Ecol_Letters.tex
Ecol Letters

Despite the astonishing diversity of floral structures among flowering plants (\protect\hyperlink{ref-barrett2002}{Barrett 2002}; \protect\hyperlink{ref-schiestl2013}{Schiestl \& Johnson 2013}) and their importance in shaping plant-pollinator interactions (\protect\hyperlink{ref-fenster2004}{Fenster \emph{et al.} 2004}; \protect\hyperlink{ref-dellinger2020}{Dellinger 2020}), a unified framework that \textbf{describes the major} plant reproductive trade-offs is currently lacking (\protect\hyperlink{ref-roddy2021}{Roddy \emph{et al.} 2021}). In addition, macroecological studies that investigate plant reproductive traits are scarce (\protect\hyperlink{ref-baude2016}{Baude \emph{et al.} 2016}; \protect\hyperlink{ref-munoz2016}{Munoz \emph{et al.} 2016}; \protect\hyperlink{ref-grossenbacher2017}{Grossenbacher \emph{et al.} 2017}; \protect\hyperlink{ref-moeller2017}{Moeller \emph{et al.} 2017}) and consequently, there is poor understanding of how reproductive traits drive interactions with pollinators at large scales (\protect\hyperlink{ref-sargent2008}{Sargent \& Ackerly 2008}; \protect\hyperlink{ref-rech2016}{Rech \emph{et al.} 2016}; \protect\hyperlink{ref-salguero2016}{Salguero-Gómez \emph{et al.} 2016}; \protect\hyperlink{ref-ruger2018}{Rüger \emph{et al.} 2018}). Linking the plant's position in trait-space with the different pollinator groups could help to improve our understanding of plant-pollinator associations (\protect\hyperlink{ref-dehling2016}{Dehling \emph{et al.} 2016}). Further, there is increasing interest in understanding drivers of plant-pollinator interactions using trait-based approaches (\protect\hyperlink{ref-fenster2004}{Fenster \emph{et al.} 2004}; \protect\hyperlink{ref-rosas2014}{Rosas-Guerrero \emph{et al.} 2014}) and trait-matching analyses (\protect\hyperlink{ref-stang2009}{Stang \emph{et al.} 2009}; \protect\hyperlink{ref-bartomeus2016}{Bartomeus \emph{et al.} 2016}). \textbf{However, plant reproductive traits have been overlooked beyond highly specialised pollination systems (\protect\hyperlink{ref-dellinger2020}{Dellinger 2020}) and remains unclear how specific plant reproductive biology traits (e.g., mating or compatibility system) influence plant-pollinator associations (\protect\hyperlink{ref-tur2013}{Tur \emph{et al.} 2013}; \protect\hyperlink{ref-devaux2014}{Devaux \emph{et al.} 2014}).}

@@

Several studies have identified links between plant traits and plant-pollinator network properties (\protect\hyperlink{ref-bartomeus2013}{Bartomeus 2013}; \protect\hyperlink{ref-olito2015}{Olito \& Fox 2015}; \protect\hyperlink{ref-rowe2020}{Rowe \emph{et al.} 2020}). Moreover, plant traits can define species' network roles (e.g., specialists vs generalists; \protect\hyperlink{ref-lazaro2013}{Lázaro \emph{et al.} 2013}; \protect\hyperlink{ref-tur2013}{Tur \emph{et al.} 2013}). For example, plant species that occupy reproductive trait space extremes are more likely to exhibit higher levels of specialisation and be more reliant on the trait-matching with pollinators (\protect\hyperlink{ref-junker2013}{Junker \emph{et al.} 2013}; \protect\hyperlink{ref-coux2016}{Coux \emph{et al.} 2016}). Morphological matching between plant and floral visitors often determines plant-pollinator interactions, and can thus strongly influence interaction network structure (\protect\hyperlink{ref-stang2009}{Stang \emph{et al.} 2009}; \protect\hyperlink{ref-ibanez2012}{Ibanez 2012}). \textbf{Importantly, the combination of traits has shown to increase the predictive power of the network interactions (\protect\hyperlink{ref-eklof2013}{Eklöf \emph{et al.} 2013}) and by considering the position in the multidimensional trait space we can determine species' functional roles within plant-pollinator networks (\protect\hyperlink{ref-dehling2016}{Dehling \emph{et al.} 2016}). Nonetheless, we know little if those patterns generally studied at the community level are representative of wider macroecological scales.}

Here, we aim to explore the potential trade-offs among reproductive traits and how these influence plant-pollinator interactions. First, we identify the major axes of reproductive trait variation and trade-offs that determine plant form and function. Second, we investigate how plant species' position in trait-space influence interactions with floral visitors. Finally, we investigate how both the main axes of trait variation, and individual traits, influence plant species' roles within networks using a set of complementary interaction network metrics (i.e., interaction strength, normalized degree and specialization).

@@

We selected 28 studies from 18 different countries that constituted a total of 64 plant-pollinator networks. These studies recorded plant-pollinator interactions in natural systems and were selected so that we had broad geographical representation. Although these studies differ in sampling effort and methodology, all studies provided information about plant-pollinator interactions (weighted and non-weighted), which we used to build a database of plant species that are likely to be animal pollinated. Many of these networks are freely available either as published studies (\protect\hyperlink{ref-olesen2007}{Olesen \emph{et al.} 2007}; \protect\hyperlink{ref-fortuna2010}{Fortuna \emph{et al.} 2010}; \protect\hyperlink{ref-carvalheiro2014}{Carvalheiro \emph{et al.} 2014}) or available in online archives (e.g., The Web of Life, \protect\hyperlink{ref-fortuna2010}{Fortuna \emph{et al.} 2010}; and Mangal, \protect\hyperlink{ref-poisot2016}{Poisot \emph{et al.} 2016}). In total, our network dataset (Table S1) constituted 60 weighted (interaction frequency) and 4 unweighted (presence/absence of the interaction) networks, each sampled at a unique location and year, as well as eight meta-webs where interactions were pooled across several locations and multiple years.

@@

\textbf{Functional traits}

@@

We explored the trade-offs between different quantitative plant functional traits with a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width respectively (Pearson's correlation = 0.72, \emph{P} \textless{} 0.01 and Pearson's correlation = 0.36, \emph{P} \textless{} 0.01), and thus we avoided overemphasizing flower size on the spectrum of trait variation. Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the association of the different qualitative traits with the main axes of trait variation. Prior to the analyses, we excluded outliers and standardized the data. Due to the high sensitivity of dimensionality reduction to outliers, we excluded values within the 2.5th--97.5th percentile range (\protect\hyperlink{ref-legendre2012}{Legendre \& Legendre 2012}), and thus our final dataset had 1,236 species. Then, we log transformed the variables to reduce the influence of outliers and z-transformed (X= 0, SD=1) so that all variables were within the same numerical range. We performed the pPCA using the function \emph{phyl.pca} from the package \emph{phytools} (\protect\hyperlink{ref-revell2012}{Revell 2012}) with the method lambda (\(\lambda\)) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode covariance because values for each variables were on the same scale following transformation (\protect\hyperlink{ref-abdi2010}{Abdi \& Williams 2010}). Moreover, to corroborate that our imputation of missing values did not affect our results, we conducted a pPCA on the full dataset without missing values (see Fig. S2). We found little difference between the explained variance with the imputed dataset (51.08\%) and the dataset without missing values (52.87\%). In addition, the loadings on each principal component had a similar contribution and correlation patterns, with the exception of plant height which showed slight variations between the imputed and non-imputed dataset. Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity. For this, we included all quantitative traits considered in the main pPCA plus pollen grains and microlitres of nectar per flower.

@@

Analyses were conducted on the subset of 60 weighted networks sampled in a unique flowering season and site, which included 556 plant and 1,126 pollinator species. These networks were analysed in their qualitative and quantitative form. First, we analysed the binary version of these weighted networks with presence-absence information that assumes equal weight across interactions. Second, we analysed the untransformed weighted networks with interaction frequency that accounts for the intensity of the interaction. Although floral visitors are not always pollinators and interaction frequency does not consider each pollinator species efficiency (\protect\hyperlink{ref-ballantyne2015}{Ballantyne \emph{et al.} 2015}), interaction frequency can provide valuable information of the contribution of floral visitors to pollination (\protect\hyperlink{ref-vazquez2005}{Vázquez \emph{et al.} 2005}, \protect\hyperlink{ref-vazquez2012}{2012}). In total, our network dataset (excluding meta-webs and non-weighted networks) included 2,256 interactions of bees with plants, 1,768 non-syrphid-Diptera interactions, 845 syrphids interactions, 437 Lepidoptera interactions, 432 Coleoptera interactions and 362 non-bee-Hymenoptera interactions. Sampling methods varied across networks but this was accounted for in analyses by considering them in the random effects of the modelling process. All analyses were conducted in R version \emph{4.0.3}.

@@

We used Bayesian modelling (see below for details) to explore the effect of floral visitor groups and the main axes of trait variation (pPCA with imputed dataset) on both qualitative and quantitative floral interactions per plant species. For this, we divided floral visitors into six main guilds that differ in life form, behaviour and are likely to play a similar ecological role: (i) bees (Hymenoptera-Anthophila), (ii) non-bee-Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Diptera-Syrphidae), (iv) non-syrphid-Diptera (Diptera-non-Syrphidae), (v) Lepidoptera and (vi) Coleoptera. Moreover, because the guild of bees was the most represented group with 2,256 records and had the highest frequency of visits, we also explored the presence-absence of interaction and visitation rate of the main bee families (Andrenidae, Apidae, Colletidae, Halictidae and Megachilidae) on the trait space. In addition, we found that \emph{Apis mellifera} was the floral visitor with the largest proportion of records counted (7.55\% of the total). This finding is consistent with previous research showing that \emph{A. mellifera} was the most frequent floral visitor in a similar dataset of 80 plant-pollinator networks in natural ecosystems (\protect\hyperlink{ref-hung2018}{Hung \emph{et al.} 2018}). Hence, to control for the effect of \emph{A. mellifera} on the observed visitation patterns of bees, we conducted an analogous analysis with presence-absence of interaction and visitation rate excluding \emph{A. mellifera}. We found that \emph{A. mellifera}, was partly driving some of the observed trends on PC1 (Fig. S3). However, we did not detect major differences on PC2 and PC3.

We implemented Bayesian generalized linear mixed models using the R package \emph{brms} (\protect\hyperlink{ref-burkner2017}{Bürkner 2017}). We modelled the frequency of visits as a function of the main axes of plant trait variation and their interactions with floral visitor functional groups (Visits \textasciitilde{} PC1 x FGs + PC2 x FGs + PC3 x FGs). Because we were interested in possible differences in the visitation patterns among floral visitors groups to plants with different strategies, we included interactions between the main axes of trait variation (PC1, PC2 and PC3) and the floral visitor guilds. In this model, we added a nested random effect of networks nested within the study system to capture the variation in networks among studies and within networks. Moreover, we included the phylogenetic covariance matrix as a random factor due to the possible shared evolutionary histories of species and therefore lack of independence across them. We specified this model with a zero inflated negative binomial distribution and weakly informative priors from the brms function. We run this model for 3,000 iterations and with previous 1,000 warm up iterations. We set delta (\(\Delta\)) to 0.99 to avoid divergent transitions and visualized the posterior predictive checks with the function \emph{pp\_check} using the \emph{bayesplot} package (\protect\hyperlink{ref-gabry2019}{Gabry \emph{et al.} 2019}).

\textbf{Plant species functional roles}

We investigated whether different quantitative traits determined plant species functional roles using Bayesian modelling and regression trees. For this, we selected simple and complementary species-level network metrics commonly applied in bipartite network studies (\protect\hyperlink{ref-dormann2008}{Dormann \emph{et al.} 2008}) with a straightforward ecological interpretation relevant to our research goals. The different plant species-level metrics were: (i) sum of visits per plant species; (ii) normalized degree, calculated as the number of links per plant species divided by the total possible number of partners; and (iii) specialization (d') (\protect\hyperlink{ref-bluthgen2006}{Blüthgen \emph{et al.} 2006}), which measures the deviation of an expected random choice of the available interaction partners and ranges between 0 (maximum generalization) and 1 (maximum specialization). Normalized degree and specialization were calculated with the \emph{specieslevel} function from the R package \emph{bipartite} (\protect\hyperlink{ref-dormann2008}{Dormann \emph{et al.} 2008}).

First, we modelled the distinct plant species metrics (sum of visits, normalized degree and plant specialization) as a function of the three main axes of trait variation (plant species level metric \textasciitilde{} PC1 + PC2 + PC3). For each response variable (i.e., each plant species level metric), we used different distribution families (zero inflated negative binomial for the sum of visits, weibull for normalized degree and zero one inflated beta for specialization). Finally, we used the same random factors, model settings and conducted the same posterior predictive checks for each model as detailed above in the `visitation patterns section.'

Second, to better understand these complex trait relationships, we used regression trees. Regression trees are recursive algorithms which can detect complex relationships among predictors and allow identification of the relevance of specific trait combinations on species functional roles. We focused exclusively on quantitative traits because almost all categorical traits were statistically associated with the first two axes of trait variation (Table S2). We conducted this analysis using the \emph{rpart} package (\protect\hyperlink{ref-therneau2015}{Therneau \emph{et al.} 2015}) with method \emph{`anova'} with a minimum of 50 observations per terminal node and we used the \emph{rpart.plot} package (\protect\hyperlink{ref-milborrow2015}{Milborrow 2015}) to plot the regression trees. We considered the species level indices as response variables (interaction frequency, normalized degree and specialization) and we performed one regression tree per metric using the different quantitative traits as predictors. We calculated two regression trees per plant species-level metric, one for the full set of species and another for the subset of species for which we had pollen and nectar traits. We focused on regression trees that included floral rewards because they consistently showed pollen and nectar traits as being the best for explaining the different species-level metrics (Fig. S4).

@@

The main axes of trait variation explained partly presence-absence interactions between plant and floral visitors (conditional \(R^{2}\) = 0.26; marginal \(R^{2}\) = 0.20) but little of the overall visitation rates (conditional \(R^{2}\) = 0.31; marginal \(R^{2}\) = 0.06). However, we found relevant trends across the different floral visitor guilds on both presence-absence and visitation interactions (Fig. 3). On the pollinator dependence trade-off, all floral visitor guilds interacted more frequently with plant species with higher pollinator dependence (PC2; Fig. 3b and Fig. 3e). For presence-absence interactions we found that all Diptera, Coleoptera and non-bee-Hymenoptera guilds interacted more frequently with plants with high flower number and small flowers (flower number - flower size trade-off, PC1; Fig. 3a) but bees and Lepidoptera interacted slightly more frequently with plant species with low flower number but large flowers. For presence-absence interactions on PC3 (style length trade-off; Fig. 3c), we found that bees interacted clearly more with plant species with long styles and high selfing and the rest of the guilds interacted slightly more with plant species with short styles and low selfing. In addition, all guilds other than Syrphids and Lepidoptera (i.e., all Hymenoptera, non-syrphid-Diptera and Coleoptera) showed greater visitation rates on species with small numerous flowers (PC1; Fig. 3d). On the style length trade-off, bees, Lepidoptera and non-bee-Hymenoptera showed greater visitation rates on plant species with larger styles and higher levels of selfing; while syrphids, non-syrphid-Diptera and Coleoptera showed higher visitation rates on species with shorter styles and lower selfing (Fig. 3f).

The additional model for both presence-absence of interactions (marginal \(R^{2}\) = 0.29; conditional \(R^{2}\) = 0.19) and visitation rate (marginal \(R^{2}\) = 0.30; conditional \(R^{2}\) = 0.03) for the most represented families of bees showed that the family Apidae was the main driver of the observed patterns. The contrasting differences between presence-absence and visitation rate for bees on PC1 (Fig. 3a and Fig. 3d) were driven by the family Andrenidae, which interacted more frequently on presence-absence interactions with plant species with low number of large flowers (Fig. S8).

@@

\begin{figure}
\centering
\includegraphics{output/figures/unnamed-chunk-4-1.pdf}
\caption{\label{fig:unnamed-chunk-4}\textbf{Figure 3} Interaction (yes/no) and visitation rates across the three main axes of trait variation per floral visitor guild. Fitted posterior estimates of the presence/absence of interaction (a, b and c) and number of visits (d, e and f) made by the different floral visitors guilds in relation to PC1, PC2 and PC3. PC1 represents the flower number - flower size trade-off, PC2 represents the pollinator dependence trade-off and PC3, the style length trade-off. For visualization purposes, due to large differences between the visitation rates of bees and the rest of guilds, the number of visits was log-transformed (Y-axis of lower panel).}
\end{figure}

@@

\textbf{Plant species functional roles}

The variance of the different plant species-level network metrics was poorly explained by the three main axes of trait variation (Fig. S9; interaction frequency \textasciitilde{} PCs, conditional \(R^{2}\) = 0.11, marginal \(R^{2}\) = 0.02; normalized degree \textasciitilde{} PCs, conditional \(R^{2}\) = 0.24, marginal \(R^{2}\) = 0.02; and, specialization \textasciitilde{} PCs, conditional \(R^{2}\) = 0.37, marginal\(R^{2}\) = 0.03). Overall, the most notable trends were found on PC1 and PC3 for interaction frequency and specialization. On the flower number - flower size trade-off (PC1), interaction frequency was higher for plant species with more flowers but was lower for plant species with larger flowers (Fig. S9a). On PC1, specialization showed the opposite trend (Fig. S9g). On the style length trade-off (PC3), interaction frequency was lower for plants with shorter styles and lower autonomous selfing and higher for species with longer styles and higher autonomous selfing (Fig. S9c). Again, specialization showed the opposite trend to interaction frequency (Fig. S9i).

>>>>

Despite the astonishing diversity of floral structures among flowering plants (\protect\hyperlink{ref-barrett2002}{Barrett 2002}; \protect\hyperlink{ref-schiestl2013}{Schiestl \& Johnson 2013}) and their importance in shaping plant-pollinator interactions (\protect\hyperlink{ref-fenster2004}{Fenster \emph{et al.} 2004}; \protect\hyperlink{ref-dellinger2020}{Dellinger 2020}), a unified framework that describes the major plant reproductive trade-offs is currently lacking (\protect\hyperlink{ref-roddy2021}{Roddy \emph{et al.} 2021}). In addition, macroecological studies that investigate plant reproductive traits are scarce (\protect\hyperlink{ref-baude2016}{Baude \emph{et al.} 2016}; \protect\hyperlink{ref-munoz2016}{Munoz \emph{et al.} 2016}; \protect\hyperlink{ref-grossenbacher2017}{Grossenbacher \emph{et al.} 2017}; \protect\hyperlink{ref-moeller2017}{Moeller \emph{et al.} 2017}) and consequently, there is poor understanding of how reproductive traits drive interactions with pollinators at large scales (\protect\hyperlink{ref-sargent2008}{Sargent \& Ackerly 2008}; \protect\hyperlink{ref-rech2016}{Rech \emph{et al.} 2016}; \protect\hyperlink{ref-salguero2016}{Salguero-Gómez \emph{et al.} 2016}; \protect\hyperlink{ref-ruger2018}{Rüger \emph{et al.} 2018}). Linking the plant's position in trait-space with the different pollinator groups could help to improve our understanding of plant-pollinator associations (\protect\hyperlink{ref-dehling2016}{Dehling \emph{et al.} 2016}). Further, there is increasing interest in understanding drivers of plant-pollinator interactions using trait-based approaches (\protect\hyperlink{ref-fenster2004}{Fenster \emph{et al.} 2004}; \protect\hyperlink{ref-rosas2014}{Rosas-Guerrero \emph{et al.} 2014}) and trait-matching analyses (\protect\hyperlink{ref-stang2009}{Stang \emph{et al.} 2009}; \protect\hyperlink{ref-bartomeus2016}{Bartomeus \emph{et al.} 2016}). However, plant reproductive traits have been overlooked beyond highly specialised pollination systems (\protect\hyperlink{ref-dellinger2020}{Dellinger 2020}) and remains unclear how specific plant reproductive biology traits (e.g., mating or compatibility system) influence plant-pollinator associations (\protect\hyperlink{ref-tur2013}{Tur \emph{et al.} 2013}; \protect\hyperlink{ref-devaux2014}{Devaux \emph{et al.} 2014}).

@@

Several studies have identified links between plant traits and plant-pollinator network properties (\protect\hyperlink{ref-bartomeus2013}{Bartomeus 2013}; \protect\hyperlink{ref-olito2015}{Olito \& Fox 2015}; \protect\hyperlink{ref-rowe2020}{Rowe \emph{et al.} 2020}). Moreover, plant traits can define species' network roles (e.g., specialists vs generalists; \protect\hyperlink{ref-lazaro2013}{Lázaro \emph{et al.} 2013}; \protect\hyperlink{ref-tur2013}{Tur \emph{et al.} 2013}). For example, plant species that occupy reproductive trait space extremes are more likely to exhibit higher levels of specialisation and be more reliant on the trait-matching with pollinators (\protect\hyperlink{ref-junker2013}{Junker \emph{et al.} 2013}; \protect\hyperlink{ref-coux2016}{Coux \emph{et al.} 2016}). Morphological matching between plant and floral visitors often determines plant-pollinator interactions, and can thus strongly influence interaction network structure (\protect\hyperlink{ref-stang2009}{Stang \emph{et al.} 2009}; \protect\hyperlink{ref-ibanez2012}{Ibanez 2012}). Importantly, the combination of traits has shown to increase the predictive power of the network interactions (\protect\hyperlink{ref-eklof2013}{Eklöf \emph{et al.} 2013}) and by considering the position in the multidimensional trait space we can determine species' roles within plant-pollinator interaction networks (\protect\hyperlink{ref-dehling2016}{Dehling \emph{et al.} 2016}). Nonetheless, we know little if those patterns generally studied at the community level are representative of wider macroecological scales.

Here, we aim to explore the potential trade-offs among reproductive traits and how these influence plant-pollinator interactions. First, we identify the major axes of reproductive trait variation and trade-offs that determine plant form and function. Second, we investigate how plant species' position in trait-space influence interactions with floral visitors. Finally, we investigate how both the main axes of trait variation, and individual traits, influence plant species' roles within networks using a set of complementary interaction network metrics (i.e., visitation rate, normalized degree and specialization).

@@

We selected 28 studies from 18 different countries that constituted a total of 64 plant-pollinator networks. These studies recorded plant-pollinator interactions in natural systems and were selected so that we had broad geographical representation. Although these studies differ in sampling effort and methodology, all studies provided information about plant-pollinator interactions (weighted and non-weighted), which we used to build a database of plant species that are likely to be animal pollinated. Many of these networks are freely available either as published studies (\protect\hyperlink{ref-olesen2007}{Olesen \emph{et al.} 2007}; \protect\hyperlink{ref-fortuna2010}{Fortuna \emph{et al.} 2010}; \protect\hyperlink{ref-carvalheiro2014}{Carvalheiro \emph{et al.} 2014}) or available in online archives (e.g., The Web of Life, \protect\hyperlink{ref-fortuna2010}{Fortuna \emph{et al.} 2010}; and Mangal, \protect\hyperlink{ref-poisot2016}{Poisot \emph{et al.} 2016}). In total, our network dataset (Table S1) constituted 60 weighted (visitation rate) and 4 unweighted (presence-absence of the interaction) networks, each sampled at a unique location and year, as well as eight meta-webs where interactions were pooled across several locations and multiple years.

@@

\textbf{Plant traits}

@@

We explored the trade-offs between the different quantitative plant traits with a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width respectively (Pearson's correlation = 0.72, \emph{P} \textless{} 0.01 and Pearson's correlation = 0.36, \emph{P} \textless{} 0.01), and thus we avoided overemphasizing flower size on the spectrum of trait variation. Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the association of the different qualitative traits with the main axes of trait variation. Prior to the analyses, we excluded outliers and standardized the data. Due to the high sensitivity of dimensionality reduction to outliers, we excluded values within the 2.5th--97.5th percentile range (\protect\hyperlink{ref-legendre2012}{Legendre \& Legendre 2012}), and thus our final dataset had 1,236 species. Then, we log transformed the variables to reduce the influence of outliers and z-transformed (X= 0, SD=1) so that all variables were within the same numerical range. We performed the pPCA using the function \emph{phyl.pca} from the package \emph{phytools} (\protect\hyperlink{ref-revell2012}{Revell 2012}) with the method lambda (\(\lambda\)) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode covariance because values for each variables were on the same scale following transformation (\protect\hyperlink{ref-abdi2010}{Abdi \& Williams 2010}). Moreover, to corroborate that our imputation of missing values did not affect our results, we conducted a pPCA on the full dataset without missing values (Fig. S2). We found little difference between the explained variance with the imputed dataset (51.08\%) and the dataset without missing values (52.87\%). In addition, the loadings on each principal component had a similar contribution and correlation patterns, with the exception of plant height which showed slight variations between the imputed and non-imputed dataset. Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity. For this, we included all quantitative traits considered in the main pPCA plus pollen grains and microlitres of nectar per flower.

@@

Analyses were conducted on the subset of 60 weighted networks sampled in a unique flowering season and site, which included 556 plant and 1,126 pollinator species. These networks were analysed in their qualitative and quantitative form. First, we analysed the binary version of these weighted networks with presence-absence information that assumes equal weight across interactions. Second, we analysed the untransformed weighted networks with visitation rate that accounts for the intensity of the interaction. Although floral visitors are not always pollinators and visitation rate does not consider each pollinator species efficiency (\protect\hyperlink{ref-ballantyne2015}{Ballantyne \emph{et al.} 2015}), visitation rate can provide valuable information of the contribution of floral visitors to pollination (\protect\hyperlink{ref-vazquez2005}{Vázquez \emph{et al.} 2005}, \protect\hyperlink{ref-vazquez2012}{2012}). In total, our network dataset (excluding meta-webs and non-weighted networks) included 2,256 interactions of bees with plants, 1,768 non-syrphid-Diptera interactions, 845 syrphids interactions, 437 Lepidoptera interactions, 432 Coleoptera interactions and 362 non-bee-Hymenoptera interactions. Sampling methods varied across networks but this was accounted for in analyses by considering them in the random effects of the modelling process. All analyses were conducted in R version \emph{4.0.3}.

@@

We used Bayesian modelling (see below for details) to explore the effect of floral visitor groups and the main axes of trait variation (pPCA with imputed dataset) on both qualitative and quantitative floral interactions per plant species. For this, we divided floral visitors into six main guilds that differ in life form, behaviour and are likely to play a similar ecological role: (i) bees (Hymenoptera-Anthophila), (ii) non-bee-Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Diptera-Syrphidae), (iv) non-syrphid-Diptera (Diptera-non-Syrphidae), (v) Lepidoptera and (vi) Coleoptera. Moreover, because the guild of bees was the most represented group with 2,256 records and had the highest frequency of visits, we also explored presence-absence of the interaction and visitation rate of the main bee families (Andrenidae, Apidae, Colletidae, Halictidae and Megachilidae) on the trait space. In addition, we found that \emph{Apis mellifera} was the floral visitor with the largest proportion of records counted (7.55\% of the total). This finding is consistent with previous research showing that \emph{A. mellifera} was the most frequent floral visitor in a similar dataset of 80 plant-pollinator networks in natural ecosystems (\protect\hyperlink{ref-hung2018}{Hung \emph{et al.} 2018}). Hence, to control for the effect of \emph{A. mellifera} on the observed visitation patterns of bees, we conducted an analogous analysis with presence-absence of the interaction and visitation rate excluding \emph{A. mellifera}. We found that \emph{A. mellifera}, was partly driving some of the observed trends on PC1 (Fig. S3). However, we did not detect major differences on PC2 and PC3.

We implemented Bayesian generalized linear mixed models using the R package \emph{brms} (\protect\hyperlink{ref-burkner2017}{Bürkner 2017}). We modelled presence-absence of observed interactions and visitation rate as a function of the main axes of plant trait variation and their interactions with floral visitor guilds (e.g., visitation rate \textasciitilde{} PC1 x FGs + PC2 x FGs + PC3 x FGs). Because we were interested in possible differences in the visitation patterns among floral visitors groups to plants with different strategies, we included interactions between the main axes of trait variation (PC1, PC2 and PC3) and the different floral visitor guilds. We added a nested random effect of networks nested within the study system to capture the variation in networks among studies and within networks. Moreover, we included the phylogenetic covariance matrix as a random factor due to the possible shared evolutionary histories of species and therefore lack of independence across them. We specified for presence-absence of interaction and visitation rate a Bernoulli and a zero inflated negative binomial distribution, respectively. The models were run with weakly informative priors from the brms function, 3,000 iterations and with previous 1,000 warm up iterations. We set delta (\(\Delta\)) to 0.99 to avoid divergent transitions and visualized the posterior predictive checks with the function \emph{pp\_check} using the \emph{bayesplot} package (\protect\hyperlink{ref-gabry2019}{Gabry \emph{et al.} 2019}).

\textbf{Plant species network roles}

We investigated whether different quantitative traits determined plant species network roles using Bayesian modelling and regression trees. For this, we selected simple and complementary species-level network metrics commonly applied in bipartite network studies (\protect\hyperlink{ref-dormann2008}{Dormann \emph{et al.} 2008}) with a straightforward ecological interpretation relevant to our research goals. The different plant species-level metrics were: (i) sum of visits per plant species; (ii) normalized degree, calculated as the number of links per plant species divided by the total possible number of partners; and (iii) specialization (d') (\protect\hyperlink{ref-bluthgen2006}{Blüthgen \emph{et al.} 2006}), which measures the deviation of an expected random choice of the available interaction partners and ranges between 0 (maximum generalization) and 1 (maximum specialization). Normalized degree and specialization were calculated with the \emph{specieslevel} function from the R package \emph{bipartite} (\protect\hyperlink{ref-dormann2008}{Dormann \emph{et al.} 2008}).

First, we modelled the distinct plant species metrics (sum of visits, normalized degree and plant specialization) as a function of the three main axes of trait variation (plant species level metric \textasciitilde{} PC1 + PC2 + PC3). For each response variable (i.e., each plant species level metric), we used different distribution families (zero inflated negative binomial for the sum of visits, weibull for normalized degree and zero one inflated beta for specialization). Finally, we used the same random factors, model settings and conducted the same posterior predictive checks for each model as detailed above in the `visitation patterns' section.

Second, to better understand complex trait relationships, we used regression trees. Regression trees are recursive algorithms which can detect complex relationships among predictors and allow identification of the relevance of specific trait combinations on explaining species roles within the network of interaction. We focused exclusively on quantitative traits because almost all categorical traits were statistically associated with the first two axes of trait variation (Table S2). We conducted this analysis using the \emph{rpart} package (\protect\hyperlink{ref-therneau2015}{Therneau \emph{et al.} 2015}) with method \emph{`anova'} with a minimum of 50 observations per terminal node and we used the \emph{rpart.plot} package (\protect\hyperlink{ref-milborrow2015}{Milborrow 2015}) to plot the regression trees. We considered the species level indices as response variables (visitation rate, normalized degree and specialization) and we performed one regression tree per metric using the different quantitative traits as predictors. We calculated two regression trees per plant species-level metric, one for the full set of species and another for the subset of species for which we had pollen and nectar traits. We focused on regression trees that included floral rewards because they consistently showed pollen and nectar traits as being the best for explaining the different species-level metrics (Fig. S4).

@@

The main axes of trait variation explained partly presence-absence of interaction partners (conditional \(R^{2}\) = 0.26; marginal \(R^{2}\) = 0.20) but little of the overall visitation rates (conditional \(R^{2}\) = 0.31; marginal \(R^{2}\) = 0.06). However, we found relevant differences across the different floral visitor guilds on both presence-absence of interactions and visitation rates (Fig. 3). We found that plants with high flower number and small flowers had higher interaction partners of Coleoptera, non-bee-Hymenoptera and all Diptera guilds but plants with low flower number but large flowers had higher interaction partners of bees and Lepidoptera guilds (flower number - flower size trade-off, PC1; Fig. 3a). However, plant species with high flower number but small flowers had higher visitation rates of bees and syrphids guilds (PC1; Fig. 3d). Remarkably, all plant species with higher pollinator dependence had higher number of interacting partners and visitation rates for all floral visitor guilds (PC2; Fig. 3b and Fig. 3e). Finally, plant species with short styles and low selfing had higher interaction partners of all guilds but bees that interacted clearly more with plant species with long styles and high selfing (style length trade-off; Fig. 3c). However, for visitation rate, we found that plants with long styles and high selfing interacted more frequently with Lepidoptera and non-bee-Hymenoptera guilds (Fig. 3f).

The additional model for both presence-absence of interaction (marginal \(R^{2}\) = 0.29; conditional \(R^{2}\) = 0.19) and visitation rate (marginal \(R^{2}\) = 0.30; conditional \(R^{2}\) = 0.03) for the most represented families of bees showed that the family Apidae was the main driver of the observed patterns. The contrasting differences between presence-absence of interaction and visitation rate for bees on PC1 (Fig. 3a and Fig. 3d) were driven by the family Andrenidae that had higher number of interacting partners but lower visitation rates on plant species with low number of large flowers (Fig. S8).

@@

\begin{figure}
\centering
\includegraphics{output/figures/unnamed-chunk-4-1.pdf}
\caption{\label{fig:unnamed-chunk-4}\textbf{Figure 3} Fitted posterior estimates of the presence-absence of interaction (a, b and c) and visitation rate (d, e and f) of the different floral visitor guilds in relation to the main axes of trait variation (PC1, PC2 and PC3). PC1 represents the flower number - flower size trade-off, PC2 represents the pollinator dependence trade-off and PC3, the style length trade-off. For visualization purposes, due to large differences between the visitation rates of bees and the rest of guilds, the number of visits was log-transformed (Y-axis of lower panel).}
\end{figure}

@@

\textbf{Plant species network roles}

The variance of the different plant species-level network metrics was poorly explained by the three main axes of trait variation (Fig. S9; visitation rate \textasciitilde{} PCs, conditional \(R^{2}\) = 0.11, marginal \(R^{2}\) = 0.02; normalized degree \textasciitilde{} PCs, conditional \(R^{2}\) = 0.24, marginal \(R^{2}\) = 0.02; and, specialization \textasciitilde{} PCs, conditional \(R^{2}\) = 0.37, marginal\(R^{2}\) = 0.03). Overall, the most notable trends were found on PC1 and PC3 for visitation rate and specialization. On the flower number - flower size trade-off (PC1), visitation rate was higher for plant species with more flowers but was lower for plant species with larger flowers (Fig. S9a). On PC1, specialization showed the opposite trend (Fig. S9g). On the style length trade-off (PC3), visitation rate was lower for plants with shorter styles and lower autonomous selfing and higher for species with longer styles and higher autonomous selfing (Fig. S9c). Again, specialization showed the opposite trend to visitation rate (Fig. S9i).
